# Results
```{r}
#All Imports here
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(choroplethr)
library(tidyverse)
library(mapview)
library(lubridate)
library(geofacet)
library(plotly)
options(warn=-1)
```

```{r}
#Preprocessing and converting to minifiles that will be used throughout to plot
data = read.csv('/Users/vritansh/Documents/Columbia/EDAV/EDAV_FINAL_PROJECT/USChronicDiseases.csv')

```


```{r}
#We clean the dataset for plots 
clean_dataset <- function(data) {
#Get only useful columns, Drop the ones not required
data <-dplyr::select(data, -c('StratificationCategoryID2', 'StratificationCategoryID3','StratificationID2','StratificationID3', 'StratificationCategory2','Stratification2','StratificationCategory3','Stratification3','Response','LowConfidenceLimit', 'HighConfidenceLimit', 'ResponseID', 'DataValueUnit', 'DataValueAlt'))
          
          #Consider data only for 2018-2020
          df <- data[data$YearStart %in% c('2018', '2019', '2020'), ]
          
          #Select only number
          df <- df[df$DataValueType %in% c('Number'), ]
          
          #Preprocess data here 
          df$DataValue[df$DataValue==""] <- 0
          
          #replace data with a numerical value
          df$DataValue <- as.numeric(df$DataValue)
          
          #Add code for NVSS
          df <- df[df$DataSource %in% c('NVSS'), ]

          #Not required as we're not dealing with lat long values currently    
          #Provide reference for the X and Y 
          df[c("Longitude", "Latitude")] <- do.call(rbind, lapply(strsplit(df$GeoLocation , "[()]"), function(col) {
            (parts <- unlist(strsplit(col[2], " ")))
          }))
          
          #Longitude
          df$Longitude<-as.double(df$Longitude)
          options(digits=16)
          
          #Latitude
          df$Latitude<-as.double(df$Latitude)
          options(digits=16)
          
          #Remove Missing Values from Latitude and Longitude
          df<-df[!is.na(df$Latitude), ]
          
          df<-df[!is.na(df$Longitude), ]
          
          return(df)
}

df <- clean_dataset(data = data)
write.csv(df, "eda_geo_spliced.csv", row.names=FALSE)

```

## Topic Distribution 

```{r}
#Using the df data above
data <- read.csv('https://raw.githubusercontent.com/vritansh/geochronicdisease/main/data/eda_geo_spliced.csv')
```

```{r}
data %>%
  group_by(Topic) %>%
  summarise(count = sum(DataValue)) %>%
  ggplot(aes(x=fct_reorder(Topic, count,.desc = FALSE),y=count)) +
  geom_bar(stat = "identity",fill = "purple") +
  ggtitle("Topic Distribution") +
  coord_flip()+
  xlab("") +
  theme(panel.grid.major.x = element_blank())
```

The three diseases that we are considering are "Cardiovascular Disease", "Chronic Obstructive Pulmonary Disease" and "Diabetes".

```{r}
data <- data[data$Topic %in% c("Cardiovascular Disease", "Chronic Obstructive Pulmonary Disease", "Diabetes", "Overarching Conditions", "Chronic Kidney Disease"), ]
dim(data)
``` 

## Gender Distribution

```{r}
data_distribution <- data %>% group_by(Topic, StratificationCategoryID1, Stratification1) %>% summarise(total_count=sum(DataValue),  .groups ='drop')

```

```{r}
data_G <- data_distribution[data_distribution$StratificationCategoryID1 %in% c("GENDER"), ]
```

```{r}
ggplot(data_G, aes(x = Stratification1, y = total_count)) +
  geom_col(position = "dodge", fill = "green") +
  ggtitle("Gender Distribution") +
  xlab("") +
  ylab("Count")+
  facet_wrap(~Topic)
```

## Graph Showing the Top Cardiovascular Diseases


```{r}
df<- read.csv('https://raw.githubusercontent.com/vritansh/geochronicdisease/main/data/eda_geo_spliced.csv')

plot_geospatial_us <- function(df_disease, title, legend) {
  
  #Aggregate all values for a particular state
  df_disease <- df_disease %>% group_by(LocationDesc) %>% summarise(total_count=sum(DataValue),  .groups='drop') 
  
  #Create state and value column names for cholorpleth
  colnames(df_disease) <- c('state','value')
  
  #Preprocess data for graph
  df_disease_display <- df_disease  %>% 
    as.data.frame() %>% 
    transmute(region = tolower(`state`), value = df_disease$value)
  
  #Display choropleth
  state_choropleth(df_disease_display,
                   title = title,
                   legend = legend)
}

```

## Geomap : Mortality due to Cardiovascular Diseases
The plot shows the total count of the cardiovascular diseases across multiple states. The data belongs to the below topics and has been aggregated to provide complete picture.

Questions
[1] "Mortality from coronary heart disease"          
[2] "Mortality from total cardiovascular diseases"   
[3] "Mortality from cerebrovascular disease (stroke)"
[4] "Mortality from diseases of the heart"       
[5] "Mortality from heart failure"

States with highest mortality 
Texas, New York, Illinois, California, Florida 

States with lowest mortality 
Montana, North Dakota, Wyoming, South Dakota, Utah


```{r}
#"Cardiovascular Disease", "Chronic Obstructive Pulmonary Disease" and "Diabetes"
df_disease <- df[df$Topic == 'Cardiovascular Disease', ]
plot_geospatial_us(df_disease = df_disease,title = "Mortality due to Cardiovascular disease", legend = "Count of Mortality")
```


## Geomap : Mortality due to Chronic Obstructive Pulmonary Disease
The plot shows the total count of the cardiovascular diseases across multiple states. The data belongs to the below topics and has been aggregated to provide complete picture.

Questions
[1] "Mortality from coronary heart disease"          
[2] "Mortality from total cardiovascular diseases"   
[3] "Mortality from cerebrovascular disease (stroke)"
[4] "Mortality from diseases of the heart"       
[5] "Mortality from heart failure"

States with highest mortality 
Texas, New York, Ohio, California, Florida 

States with lowest mortality 
North Dakota, Wyoming, South Dakota, Utah, New Mexico


```{r}
#"Cardiovascular Disease", "Chronic Obstructive Pulmonary Disease" and "Diabetes"
df_disease <- df[df$Topic == 'Chronic Obstructive Pulmonary Disease', ]
plot_geospatial_us(df_disease = df_disease,title = "Mortality due to Chronic Obstructive Pulmonary Disease", legend = "Count of Mortality")
```




## Geomap : Mortality due to Diabetes
The plot shows the total count of the cardiovascular diseases across multiple states. The data belongs to the below topics and has been aggregated to provide complete picture.

Questions
[1] "Mortality with diabetic ketoacidosis reported as any listed cause of death"
[2] "Mortality due to diabetes reported as any listed cause of death" 

States with highest mortality 
Texas, New York, Ohio, California, Florida 

States with lowest mortality 
Montana, North Dakota, Wyoming, South Dakota, Utah


```{r}
#"Cardiovascular Disease", "Chronic Obstructive Pulmonary Disease" and "Diabetes"
df_disease <- df[df$Topic == 'Diabetes', ]
plot_geospatial_us(df_disease = df_disease,title = "Mortality due to Chronic Obstructive Pulmonary Disease", legend = "Count of Mortality")
```


```{r}

#Extract useful information from the complete data and save it.
save_geospatial_time <- function(data){
  
df <- data[data$DataValueType %in% c('Number'), ]
#Preprocess data here 
df$DataValue[df$DataValue==""] <- 0
#Setting Data value
df$DataValue <- as.numeric(df$DataValue)
#Add code for NVSS
df <- df[df$DataSource %in% c('NVSS'), ]
#Get all states
df <- df[df$LocationAbbr %in% c('TX','CA','FL','NY'), ]
#Get all values for location abbrevations
df_spread_geo_time <- df %>% group_by(LocationAbbr,Topic, YearEnd) %>% summarise(total_count=sum(DataValue),  .groups ='drop') 
#get df spread geo time total
df_spread_geo_time <- df_spread_geo_time[order(df_spread_geo_time$total_count),]
#Save the data to a dataset
df_spread_geo_time$YearEnd <- as.Date(as.character(df_spread_geo_time$YearEnd), format = "%Y")
#Save the data to a dataset and use later on github pages.
write.csv(df_spread_geo_time, "eda_geospatial_time.csv", row.names=FALSE)
}
data = read.csv('/Users/vritansh/Documents/Columbia/EDAV/EDAV_FINAL_PROJECT/USChronicDiseases.csv')
save_geospatial_time(data)

```


## Timeseries Faceted Geoplot
In this plot we're analyzing if the mortality chronic diseases have increased in the past decade. For this we've considerered the data for one decade on the top five states from above. This would give us insights on the health care and how people have evolved. It can be helpful in predicting useful information regarding infrastructure of a state.

Key Trends 
1. Cardiovascular Disease
  Except for New York state, for all the states the mortality caused by cardiovascular disease has increased. 

2. Diabetes in Texas
  As we can observe from the graph the mortality due to diabetes in Texas was ~69,000 in the year 2017, It increased close to 116,000 in 2018 and then took a dip in 2019 to ~72,000. 
   
3. Data consistency for Diabetes Data
 As we observed in above point the data trend for diabetes data similar across other states as well. It can be accounted to the way data was collected. And this dip is only seen from 2017 to 2019. This trend is interesting as it could mean that from the years 2017 to 2019 information is either with errors or in actual these states had a dip in the mortality due to diabetes. 

```{r}

plot_geospatial_time <- function(df_spread_geo_time) {
g_timeseries_geo <- ggplot(df_spread_geo_time, aes( YearEnd , total_count, color=`Topic`)) +                                    
  geom_line(aes(group = `Topic`)) +
  scale_x_date(limits = c(as.Date("2011-02-01"), as.Date("2019-12-31")), date_breaks = "12 months", date_labels = "%Y") +
  facet_wrap(. ~ LocationAbbr, nrow = 4) +
  theme(axis.text.x = element_text(size = 8 , hjust = 1))
  ggplotly(g_timeseries_geo)
}

#Read Data From CSV that was saved in previous step from Github and then plot 
df <- read_csv('https://raw.githubusercontent.com/vritansh/geochronicdisease/main/data/eda_geospatial_time.csv')
plot_geospatial_time(df)

```


