---
title: "US_Chronic_Diseases"
author: "Maria Pratyusha"
date: "2022-12-13"
output: html_document
---

1.  Import the libraries

    ```{r}
    library(dplyr)
    library(ggplot2)
    library(forcats)
    library(tidyr)
    library(tibble)
    library(choroplethr)
    library(tidyverse)
    library(mapview)
    ```

2.  Load the dataset

    ```{r}
    data <- read.csv("ChronicDiseases.csv")
    View(data)
    ```

3.  Trim the data to only consider entries from 2018 - 2020 (COVID PERIOD)

    ```{r}
    data <- data[data$YearStart %in% c('2018', '2019', '2020'), ]
    dim(data)
    ```

4.  Remove the unnecessary columns from data

    ```{r}
    data <-dplyr::select(data, -c('StratificationCategoryID2', 'StratificationCategoryID3','StratificationID2','StratificationID3', 'StratificationCategory2','Stratification2','StratificationCategory3','Stratification3', 'LowConfidenceLimit','Response', 'HighConfidenceLimit','ResponseID','DataValueUnit','DataValueAlt','DataValueFootnoteSymbol', 'DatavalueFootnote'))
    dim(data)
    ```

5.  Since, we have many categories in Data Value, we are focusing on "Number"

    ```{r}
    data <- data[data$DataValueType %in% c("Number"), ]
    dim(data)
    ```

    ```{r}
    data$DataValue[data$DataValue==""] <- 0
    data$DataValue <- as.numeric(data$DataValue)
    str(data)
    ```

6.  Data Sources Distribution

    ```{r}
    length(unique(data$DataSource))
    ```

    There are 8 data sources contributing to the trimmed data, we will be working with the most optimal contributor.

    ```{r}
    data %>%
      group_by(DataSource) %>%
      summarise(count = n()) %>%
      ggplot(aes(x=fct_reorder(DataSource, count,.desc = FALSE),y=count)) +
      geom_bar(stat = "identity",fill = "cornflowerblue") +
      ggtitle("DataSource Distribution") +
      coord_flip()+
      xlab("") +
      theme(panel.grid.major.x = element_blank())

    ```

Since "NVSS" has most contributions, we will be using the data only from that source.

```{r}
data <- data[data$DataSource %in% c("NVSS"), ]
dim(data)
```

7.  Topic Distribution

    ```{r}

    data %>%
      group_by(Topic) %>%
      summarise(count = sum(DataValue)) %>%
      ggplot(aes(x=fct_reorder(Topic, count,.desc = FALSE),y=count)) +
      geom_bar(stat = "identity",fill = "purple") +
      ggtitle("Topic Distribution") +
      coord_flip()+
      xlab("") +
      theme(panel.grid.major.x = element_blank())
    ```

    The top three diseases are "Cardiovascular Disease", "Chronic Obstructive Pulmonary Disease" and "Diabetes".

    ```{r}
    data <- data[data$Topic %in% c("Cardiovascular Disease", "Chronic Obstructive Pulmonary Disease", "Diabetes", "Overarching Conditions", "Chronic Kidney Disease"), ]
    dim(data)
    ```

8.  Gender Distribution

```{r}
data_distribution <- data %>% group_by(Topic, StratificationCategoryID1, Stratification1) %>% summarise(total_count=sum(DataValue),  .groups ='drop')
View(data_distribution)
```

```{r}
data_G <- data_distribution[data_distribution$StratificationCategoryID1 %in% c("GENDER"), ]
View(data_G)
```

```{r}
ggplot(data_G, aes(x = Stratification1, y = total_count)) +
  geom_col(position = "dodge", fill = "green") +
  ggtitle("Gender Distribution") +
  xlab("") +
  ylab("Count")+
  facet_wrap(~Topic)
```

9.  Race Distribution

    ```{r}
    data_R <- data_distribution[data_distribution$StratificationCategoryID1 %in% c("RACE"), ]
    View(data_R)
    ```

    ```{r}
    ggplot(data_R, aes(x = Stratification1, y = total_count)) +
      geom_col(position = "dodge", fill = "green") +
      ggtitle("Race Distribution") +
      xlab("") +
      coord_flip()+
      ylab("Count")+
      facet_wrap(~Topic)
    ```

<!-- -->

10. Geospatial Analysis

    ```{r}
    plot_geospatial_us <- function(df_disease, title, legend) {
      
      #Aggregate all values for a particular state
      df_disease <- df_disease %>% group_by(LocationDesc) %>% summarise(total_count=sum(DataValue),  .groups ='drop') 
      
      #Create state and value column names for cholorpleth
      colnames(df_disease) <- c('state','value')
      
      #Plot graph
      df_disease_display <- df_disease  %>% 
        as.data.frame() %>% 
        transmute(region = tolower(`state`), value = df_disease$value)
      
      #Display choropleth
      state_choropleth(df_disease_display,
                       title = title,
                       legend = legend)
    }
    ```

    ```{r}
    df_disease <- data[data$Topic == 'Cardiovascular Disease', ]
    plot_geospatial_us(df_disease = df_disease,title = "Mortality due to Cardiovascular disease", legend = "Count of Cardiovascular Diseases")
    ```

```{r}
data_distribution_states <- data %>% group_by(Topic, StratificationCategoryID1, Stratification1) %>% summarise(total_count=sum(DataValue),  .groups ='drop')
View(data_distribution)
```

Sources - NVSS

Cleaning and T - Topic , Top 5 states

Results - Gender , Top 3 topic spread in top 5 states
